---
title: "Assignment4"
author: "Joseph Jaiyeola"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   html_document:
   toc: yes
   toc_float: yes
always_allow_html: yes
download_code: yes
---

```{r}
library(tidycensus)
library(tidyverse)
library(sf)
library(tmap)

```


# 1 Find the ACS variable for the Percent Vacant housing units for 2011 and 2019 for Harris County, TX

```{r}
v11_Profile <- load_variables(2011,
                              "acs5/profile",
                              cache = TRUE) #demographic profile tables
v19_Profile <- load_variables(2019,
                              "acs5/profile",
                              cache = TRUE) #demographic 

#Search for variables by using grep()

v11_Profile[grep(x = v11_Profile$label,
                 "Vacant housing units",
                 ignore.case = TRUE), 
            c("name", "label")]

v19_Profile[grep(x = v19_Profile$label,
                 "Vacant housing units",
                 ignore.case = TRUE),
            c("name", "label")]
```


## Extract from ACS summary file data profile variables from 2011 and 2019 for Harris County, TX Census Tracts

The data profile tables are very useful because they contain lots of pre-calculated variables.

For 2011

```{r results='hide'}
house11<-get_acs(geography = "tract",
                state="TX",
                county = "Harris County",
                year = 2011,
                variables="DP04_0003P" ,
                geometry = T,
                output = "wide")


#rename variables and filter missing cases
house11<- house11%>%
  mutate( house = DP04_0003PE,
         house_er = DP04_0003PM/1.645,
         house_cv =100* (house_er/house)) %>%
  filter(complete.cases(house), is.finite(house_cv)==T)%>%
  select(GEOID, house, house_er,house_cv)

head(house11)

```



# 2 Create a map using quantile breaks of this variable for each year

## Breaking the continuous variable into discrete bins with quantile breaks


```{r}


tm_shape(house11)+
  tm_polygons(c("house"),
              title=c("% in Vacant housing units"),
              palette="Blues",
              style="quantile",
              n=5)+
  #tm_format("World", legend.outside=T, title.size =4)+
  tm_scale_bar()+
  tm_layout(title="Harris County Vacant housing Rate Estimates in 2011- Quantile Breaks",
            title.size =1.5,
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))

```






```{r}


p1<-tm_shape(house11)+
  tm_polygons(c("house"),
              title=c("% in Vacant Housing"),
              palette="Blues",
              style="quantile",
              n=5)+
  #tm_format("World", legend.outside=T, title.size =4)+
  tm_scale_bar()+
  tm_layout(title="Harris County Vacant Housing Rate Estimates",
            title.size =1.5, 
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))


p2<-tm_shape(house11)+
  tm_polygons(c("house_cv"),
              title=c("CV Vacant Housing"),
              palette="Blues",
              style="quantile",
              n=5)+
  #tm_format("World", title="Harris County Vacant Housing Rate CV", legend.outside=T)+
  tm_layout(title="Harris County Vacant Housing Rate CV",
            title.size =1.5,
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_scale_bar()+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))



tmap_arrange(p1, p2)

```

```{r}
plot(house11$house, house11$house_cv, main = "Error in Estimates vs Estimate Size")
```

# Compare Vacant Housing rates over time


```{r, results='hide'}
house19<-get_acs(geography = "tract",
                state="TX",
                county = "Harris County",
                year = 2019,
                variables="DP04_0003P" ,
                geometry = T,
                output = "wide")


#rename variables and filter missing cases
house19<- house19%>%
  mutate( house = DP04_0003PE,
         house_er = DP04_0003PM/1.645,
         house_cv =100* (house_er/house)) %>%
  filter(complete.cases(house), is.finite(house_cv)==T)%>%
  select(GEOID, house, house_er,house_cv)

head(house19)

```


```{r}


tm_shape(house19)+
  tm_polygons(c("house"),
              title=c("% in Vacant housing units"),
              palette="Blues",
              style="quantile",
              n=5)+
  #tm_format("World", legend.outside=T, title.size =4)+
  tm_scale_bar()+
  tm_layout(title="Harris County Vacant housing Rate Estimates in 2019- Quantile Breaks",
            title.size =1.5,
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))

```


# 3 Calculate the differences between the two years, including the test for significance of the differnces

```{r}

#merge the two years worth of data
st_geometry(house19)<-NULL #strip the geometry from the 2019 data

mdat<-left_join(house11, house19, by=c("GEOID"="GEOID"))

head(mdat)
```


```{r}
acstest<-function(names,geoid, est1, err1, est2, err2, alpha, yr1, yr2, span){
  
  se1<-err1/qnorm(.90)
  se2<-err2/qnorm(.90)
  yrs1<-seq(yr1, to=yr1-span)
  yrs2<-seq(yr2, to=yr2-span)

  C<-mean(yrs2%in%yrs1)
  diff<- (est1-est2)
  test<-(est1-est2) / (sqrt(1-C)*sqrt(se1^2+se2^2))
  crit<-qnorm(1-alpha/2)
  pval<-1-pnorm(abs(test))
  result<-NULL
  result[pval > alpha]<-"insignificant change"
  result[pval < alpha & test < 0]<- "significant increase"
  result[pval < alpha & test > 0]<-"significant decrease" 
  
  data.frame(name=names,geoid=geoid, est1=est1, est2=est2, se1=se1, se2=se2,difference=diff, test=test, result=result, pval=pval)
}
```


```{r}
significance(est1=mdat$house.x,
             est2=mdat$house.y,
             moe1=mdat$house_er.x,
             moe2 = mdat$house_er.y,
             clevel = .9)
```

```{r}
mdat$signif<- significance(est1=mdat$house.x,
                           est2=mdat$house.y,
                           moe1=mdat$house_er.x,
                           moe2 = mdat$house_er.y,
                           clevel = .9)
```


```{r}

  diff1119<-acstest(names = mdat$GEOID,
                    geoid = mdat$GEOID,
                    est1 = mdat$house.x,
                    est2 = mdat$house.y,
                    err1 = mdat$house_er.x,
                    err2=mdat$house_er.y,
                    alpha = .1,
                    yr1 = 2011, yr2=2019,
                    span = 5)

head(diff1119)
table(diff1119$result)
```

Compare to tidycensus

```{r}
table(mdat$signif) #

```



# 4 Create a map showing the differences by census tract
## Make a map layout

```{r, fig.height=10, fig.width=12}
acs_merge<-left_join(mdat, diff1119, by=c("GEOID"="geoid"))

tmap_mode("plot")
p1<-tm_shape(acs_merge)+
  tm_polygons(c("house.x"),
              title=c("% in Vacant Housing in  2011"),
              palette="Blues",
              style="quantile",
              n=5)+
  #tm_format("World", legend.outside=T, title.size =4)+
  tm_scale_bar()+
  tm_layout(title="Harris County Vacant Housing Rate Estimates 2011",
            title.size =1.5,
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))

p2<-tm_shape(acs_merge)+
  tm_polygons(c("house.y"),
              title=c("% in Harris County Vacant Housing Rate Estimates in 2019"),
              palette="Blues", 
              style="quantile",
              n=5)+
  #tm_format("World", title="Harris County Vacant Housing Rate CV", legend.outside=T)+
  tm_layout(title="Harris County Vacant Housing Rate Estimate 2019",
            title.size =1.5,
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_scale_bar()+
  tm_compass()+
  tm_format("World",
            legend.position =  c("left", "bottom"),
            main.title.position =c("center"))


p3  <- tm_shape(acs_merge)+
  tm_polygons(c("result"),
              title=c("Changes"),
              palette = "Set2")+
  #tm_format("World", title="Harris County Vacant Housing Rate CV", legend.outside=T)+
  tm_layout(title="Harris County Vacant Housing Rate Estimate Changes",
            title.size =1.5, 
            legend.frame = TRUE,
            title.position = c('right', 'top'))+
  tm_scale_bar()+
  tm_compass()
  

tmap_arrange(p1, p2,  p3, nrow=2)

```


## Make an interactive map
```{r}
tmap_mode("view")
#osmtile <- tmaptools::read_osm(house11, mergeTiles = T)

#tm_shape(osmtile)+
 # tm_rgb()+
tm_shape(acs_merge)+
  tm_polygons("result",
              alpha = .7,
              title=c("Changes"),
              palette = "Set2")+
  #tm_format("World", title="Harris County Vacant Housing Rate  CV", legend.outside=T)+
  tm_layout(title="Harris County Vacant Housing Rate Estimate Changes",
            title.size =1.5)+
  tm_scale_bar()
  #tm_compass()+
  
```
